---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Amanda Sajewski"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on 
generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with 
your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a 
single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, 
agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns 
to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd()
library(tidyverse)
library(agricolae)
library(readr)
library(dplyr)
library(lubridate)
library(here)
library(htmltools)

NTL_LTER_ChemPhys_Raw <-
  read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
           stringsAsFactors = TRUE)



NTL_LTER_ChemPhys_Raw$sampledate <- as.Date(NTL_LTER_ChemPhys_Raw$sampledate,
                                            format = "%m/%d/%y")
class(NTL_LTER_ChemPhys_Raw$sampledate)

#2
mytheme<-theme_classic(base_size=14)+
  theme(axis.text= element_text(color = "black"), 
        legend.position = "top",
        legend.key.height=unit(.5,'cm'),
        legend.key.width=unit(.5,'cm'))




```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature does not change with depth
Ha: Mean lake temperature decreases with depth


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet 
the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter 
plot of temperature by depth. Add a smoothed line showing the linear model, and 
limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER_ChemPhys_Raw["Month"] <- format(as.Date
                                    (NTL_LTER_ChemPhys_Raw$sampledate), "%m")

NTL_LTR_Lake_July <- NTL_LTER_ChemPhys_Raw %>%
  filter(NTL_LTER_ChemPhys_Raw$Month=="07") %>%
  select(lakename,year4,daynum,depth,temperature_C) %>%
  na.omit()
  

#5
TempDepth <- ggplot(NTL_LTR_Lake_July,aes(x=depth, y=temperature_C))+
  geom_point()+
  ylim(0,35)+
  geom_smooth(method = lm) +
  labs(title="Temperature changes with Depth", x="Depth(m)", 
       y="Temperature (C)")
print(TempDepth)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about 
the linearity of this trend?

> Answer: Temperature and depth have a negative relationship, as depth increases temperature decreases. It is not quite a linear relationship as between 5 and
10 m temperature starts to remain steady. So a true fit line would curve and 
approach flat after 10 m.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lin_reg_NTL_LTR_Lake_July <- lm(NTL_LTR_Lake_July$temperature_C~NTL_LTR_Lake_July$depth)
summary(lin_reg_NTL_LTR_Lake_July)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which 
this finding is based, and the statistical significance of the result. Also 
mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The relationship between depth and temperature is significant. Changes in 
temperature are 73.87% explained by changes in depth. This is based on 9726 
degrees of freedom (that's how many observations are being used). It is 
statistically significant as the p-value is well under our threshold of .05. 
With the change of 1 m of depth, temperature is predicted to decrease by 1.95 degrees. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to 
explore what might the best set of predictors for lake temperature in July 
across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables(year4,daynum,depth) 
is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
LakeAIC<- lm(data=NTL_LTR_Lake_July,temperature_C ~ year4 + daynum + depth)
#calling an linear regression with all of the variables
summary(LakeAIC)
step(LakeAIC)


#10
#In the stepwise AICm for removing each variable the AIC value was higher than 
#for not removing any of the variables so according to the AIC the recommended 
#set of variables is all of them. 

Lakemodel <- lm(data=NTL_LTR_Lake_July,temperature_C ~ depth + year4 + daynum)
summary(Lakemodel)

```

11. What is the final set of explanatory variables that the AIC method suggests 
we use to predict temperature in our multiple regression? How much of the 
observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The final set of explanatory variables suggested by the AIC step wise method is depth, year4 and daynum. This explains 74.11% of the observed variance in temperature. 74.11% is marginally better than just using depth, depth explains 
73.87% of the observed variance. As the goal is to have the simplest model with 
the best fit the marginally better prediction is probably not worth adding two 
more variables.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. 
(No need to test assumptions of normality or similar variances.) Create two sets
of models: one expressed as an ANOVA models and another expressed as a linear 
model (as done in our lessons).

```{r anova.model}
#12
July_lake_anova <- aov(data=NTL_LTR_Lake_July, temperature_C ~ lakename)
summary(July_lake_anova)

July_lake_anova2 <- lm(data=NTL_LTR_Lake_July, temperature_C ~ lakename)
summary(July_lake_anova2)
#intercept: Central Long Lake
```

13. Is there a significant difference in mean temperature among the lakes? 
Report your findings. 

> Answer: 
Yes, from the anova test we see that the p-value is under our threshold of .05 
so we reject the null hypothesis that there is no difference in the means 
between sites. From the lm test we can see that the means are difference between
all the difference lakes. 


14. Create a graph that depicts temperature by depth, with a separate color for 
each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make 
your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 
degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Lake_Temp <- ggplot(NTL_LTR_Lake_July, aes(x=depth, y=temperature_C, color=lakename))+
  geom_point(alpha=.5)+
  labs(title="Lake Temperature by Depth", x="Depth (m)", y="Temperature (C)")+
  geom_smooth(method=lm,se=FALSE)+
  ylim(0,35) +
  xlim(0,15)+
  mytheme

print(Lake_Temp)
  

  
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(July_lake_anova)


```

16.From the findings above, which lakes have the same mean temperature, 
statistically speaking, as Peter Lake? Does any lake have a mean temperature 
that is statistically distinct from all the other lakes?

>Answer:
Paul Lake and Ward Lake have mean temperatures that are not statistically significantly different from the mean temperature of Peter Lake (using our 
threshold pvalue of 0.05)
No, each lake has at least one other lake with a mean temperature that is not statistically different.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test 
we might explore to see whether they have distinct mean temperatures? 

>Answer: If you wrangled the data to just look at Peter Lake and Paul Lake the 
anova test would tell you if they had distinct mean temperatures. Or you could 
use the HSD.test function to look at pairwise groupings, including just Peter
Lake and Paul Lake. A two-sample T-test would also tell you if they had distinct
temperatures.



18. Wrangle the July data to include only records for Crampton Lake and Ward 
Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean 
temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Crampton_Ward_july<- filter(NTL_LTR_Lake_July,NTL_LTR_Lake_July$lakename %in% c("Crampton Lake","Ward Lake"))

C_W_lake_test<- t.test(data=Crampton_Ward_july,temperature_C ~ lakename)
C_W_lake_test
```

>Answer: 
The t-test tells us that the alternative hypothesis is accepted that the true difference in means is not equal to zero; however, the p-value is above our 
threshold value of .05 so we accept the null hypothesis that the mean 
temperature for Ward Lake and Crampton Lake are not significantly distinct. The
p-value acceptance of the null hypothesis does match the answer for part 16 as 
the tukey test also returns that they are not statistically distinct. 
