---
title: "Assignment 8: Time Series Analysis"
author: "Amanda Sajewski"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on 
generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>`
with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each
instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a
single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r, setup}
#1
getwd()
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(scales)

mytheme<- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "side")
theme_set(mytheme)

```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data 
folder. These contain ozone concentrations at Garinger High School in North 
Carolina from 2010-2019 (the EPA air database only allows downloads for one year
at a time). Import these either individually or in bulk and then combine them 
into a single dataframe named `GaringerOzone` of 3589 observation and 20 
variables. 

```{r, message = FALSE}
#2
ozone1<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2010_raw.csv",
                  stringsAsFactors = T)
ozone2<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2011_raw.csv",
                  stringsAsFactors = T)
ozone3<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2012_raw.csv",
                  stringsAsFactors = T)
ozone4<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2013_raw.csv",
                  stringsAsFactors = T)
ozone5<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2014_raw.csv",
                  stringsAsFactors = T)
ozone6<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2015_raw.csv",
                  stringsAsFactors = T)
ozone7<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2016_raw.csv",
                  stringsAsFactors = T)
ozone8<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2017_raw.csv",
                  stringsAsFactors = T)
ozone9<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2018_raw.csv",
                  stringsAsFactors = T)
ozone10<- read.csv("./Data/Raw/Ozone_Timeseries/EPAair_O3_GaringerNC2019_raw.csv",
                   stringsAsFactors = T)

GaringerOzone <- rbind(ozone1,ozone2,ozone3,ozone4,ozone5,ozone6,ozone7,ozone8,
                       ozone9,ozone10)

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. 
We want to generate a daily dataset, so we will need to fill in any missing days
with NA. Create a new data frame that contains a sequence of dates from 
2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data 
frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of 
data frames within this function so that the final dimensions are 3652 rows and
3 columns. Call your combined data frame GaringerOzone. 

```{r Data Wrangling}
#3
GaringerOzone$Date<-as.Date(GaringerOzone$Date,format="%m/%d/%Y")
class(GaringerOzone$Date)

#4
GaringerOzone1<- GaringerOzone %>%
  select(Date,Daily.Max.8.hour.Ozone.Concentration,DAILY_AQI_VALUE)

#5
Days<- 
  as.data.frame(seq(as.Date("2010/01/01"),as.Date("2019/12/31"),"days"))
colnames(Days)[1]= "Date"
class(Days$Date)
class(GaringerOzone1$Date)

#6
GaringerOzone <- left_join(Days,GaringerOzone1,by = "Date")

#For some reason this left_join returned all NA values until I selected which 
#columns I wanted in my new data frame using a pipe function


```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we 
will plot actual concentrations in ppm, not AQI values. Format your axes
accordingly. Add a smoothed line showing any linear trend of your data. Does 
your plot suggest a trend in ozone concentration over time?

```{r Plot ozone concentration}
#7

Ozoneconc<- ggplot(GaringerOzone, aes(x=Date,
                                      y=Daily.Max.8.hour.Ozone.Concentration))+
  geom_line()+
  geom_smooth(method=lm)+
  scale_x_date(labels = date_format("%Y"),
               date_breaks = "1 year")+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
  labs( title = "Ozone over time",
        y="Ozone Concentrations (ppm)")
print(Ozoneconc)

```

>Answer:
This plot suggests that there is seasonal variation with a very slight downward
trend over time.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone 
concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r Linear Interpolation to fill in missing data}
#8
head(GaringerOzone)
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)


Garinger_Ozone_clean <- 
  GaringerOzone %>% 
  mutate(Daily.Max.8.hour.Ozone.Concentration = zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration))%>%
  mutate(DAILY_AQI_VALUE=zoo::na.approx(DAILY_AQI_VALUE))

summary(Garinger_Ozone_clean$Daily.Max.8.hour.Ozone.Concentration)

```

> Answer: We didn't use the piecewise constant b/c there is high variation 
between dates so taking the nearest neighbor as taken in piecewise constant 
would flatten that variation. Spline wouldn't be appropriate because the trend 
line is linear so using a quadratic function wouldn't fit it as well. The linear interpolation method is the most appropriate by assuming that the missing data 
fall between the last and the next measurent with a straight line drawn between 
them.


9. Create a new data frame called `GaringerOzone.monthly` that contains 
aggregated data: mean ozone concentrations for each month. In your pipe, you 
will need to first add columns for year and month to form the groupings. In a 
separate line of code, create a new Date column with each month-year combination
being set as the first day of the month (this is for graphing purposes only)

```{r Aggregating Data}
#9
GaringerOzone.monthly <- Garinger_Ozone_clean %>% 
  mutate(Month=month(Date))%>%
  mutate(Year=year(Date))%>%
  group_by(Year, Month) %>%
  summarise(Mean_ozone_concentration = 
              mean(Daily.Max.8.hour.Ozone.Concentration))



GaringerOzone.monthly <- GaringerOzone.monthly %>%
  mutate(Date=ymd(paste(Year,"-",Month,"-01")))
class(GaringerOzone.monthly$Date)
 

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` 
and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be 
sure that each specifies the correct start and end dates and the frequency of 
the time series.

```{r Time series Objects}
#10
GaringerOzone.daily.ts <- ts(Garinger_Ozone_clean$Daily.Max.8.hour.Ozone.Concentration, start = c(2010,1), frequency = 365)

GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$Mean_ozone_concentration,start=c(2010,1),frequency =12)

```


11. Decompose the daily and the monthly time series objects and plot the 
components using the `plot()` function.

```{r Decomposition of time series objects}
#11
Garinger.daily.decomposed<- stl(GaringerOzone.daily.ts,s.window = "periodic")

Garinger.mean.decomposed <- stl(GaringerOzone.monthly.ts,s.window="periodic")

plot(Garinger.daily.decomposed)

plot(Garinger.mean.decomposed)


```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r Seasonal Mann-Kendall Test}
#12
Ozone_trend_smk<- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)

Ozone_trend_smk
summary(Ozone_trend_smk)


```

> Answer: A seasonal Mann-Kendall test is the only one of the four we discussed 
that takes seasonality into account. According to our decomposed time series 
plotted, a seasonal pattern is suggested. 

13. Create a plot depicting mean monthly ozone concentrations over time, with
both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r Plots of mean monthly ozone concentrations}
#13

 Monthly_ozone_plot <-
ggplot(GaringerOzone.monthly, aes(x = Date, y = Mean_ozone_concentration)) +
  geom_point() +
  geom_line() +
  labs(title="Average Ozone Concentration 2010-2020")+
  ylab("Mean ozone concentration (ppm)") +
  geom_smooth( method = lm )+
    scale_x_date(labels = date_format("%m-%Y"),
date_breaks = "6 months") + theme(axis.text.x = element_text(angle = 85,
hjust = 1)) 

print(Monthly_ozone_plot)


```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of 
your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: The research question asked us "Have ozone concentrations changed over
the 2010s at this station?" From the graph we can see that there is slightly 
less seasonal variation as we move later into the 2010s, with lower highs. 
There is also a very slight negative trend over the decade (The seasonal Mann
Kendall test tells us that there is a minor negative trend (-0.143) with slight significance looking at a p-score not much smaller than 0.05 (p-value=0.0467))

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: 
Look at how we extracted the series components for the EnoDischarge on the 
lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare
the results with the ones obtained with the Seasonal Mann Kendall on the 
complete series.

```{r Subtracting Seasonality from the monthly Garinger Ozone data}
#15
Garinger_Ozone_nonseasonal <- as.data.frame(Garinger.mean.decomposed$time.series[,2:3])

Garinger_Ozone_nonseasonal <- mutate(Garinger_Ozone_nonseasonal,
        Observed = Garinger_Ozone_nonseasonal$trend,     
        Date = GaringerOzone.monthly$Date)
Garinger_Ozone_nonseasonal <- Garinger_Ozone_nonseasonal %>%
  select(Observed,Date)
Garinger_nonseasonal.ts <- ts(Garinger_Ozone_nonseasonal$Observed, start=c(2010,1),frequency=12)

#16

Ozone_trend_nonseasonal_smk<- Kendall::MannKendall(Garinger_nonseasonal.ts)

Ozone_trend_nonseasonal_smk
summary(Ozone_trend_nonseasonal_smk)


```

> Answer:The Mann:Kendall Test run on the monthly ozone data with seasonality extracted showed a slightly stronger negative trend, with a tau score of -0.269 compared to the seasonal data that had a tau score of -0.143. The non-seasonal
data also has a much stronger significance with a p-value far below .05 compared
to our seasonal p-value that was barely below .05 (.0467).
