---
title: "Assignment 3: Data Exploration"
author: "Amanda Sajewski"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing 
`<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each 
instruction. 
4.  Assign a useful **name to each code chunk** and include ample 
**comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a 
single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox 
in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by
manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or
`View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse,
lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for 
litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these
datasets "Neonics" and "Litter", respectively. Be sure to include the
subcommand to read strings in as factors.

```{r,  message = FALSE}

getwd()

library(tidyverse)

library(lubridate)

library(readr)
Neonics <- read.csv("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", 
                    stringsAsFactors = T)

Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
                   stringsAsFactors = T)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection 
Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. 
Neonicotinoids are a class of insecticides used widely in agriculture. 
The dataset that has been pulled includes all studies published on insects. 
Why might we be interested in the ecotoxicology of neonicotinoids on insects? 
Feel free to do a brief internet search if you feel you need more background 
information.

> Answer:Insecticides don't discriminate and neonicotinoids also harm 
pollinators which are important for continued agriculture and food production. Additionally the neonicotinoids could build up in larger animals that eat 
insects and harm them potentially leading to cascades in the food web.

3.  The Niwot Ridge litter and woody debris dataset was collected from the 
National Ecological Observatory Network, which collectively includes 81 aquatic 
and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample 
forest litter and woody debris, and we will focus on the Niwot Ridge long-term 
ecological research (LTER) station in Colorado. Why might we be interested in 
studying litter and woody debris that falls to the ground in forests? Feel free 
to do a brief internet search if you feel you need more background information.

> Answer: Woody debris could serve as tinder during forest fires. It also serves 
as nutrient source for downstream aquatic ecosystems and as a habitat for 
insects and small animals on the forest floor.Leaf litter will protect the soil 
from drying out and eventually add more humus and nutrients to the soil for 
growing plants.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of 
salient information about the sampling methods here:

> Answer: 
 1.Ground traps are sampled once per year. Target sampling frequency for 
 elevated traps varies by vegetation present at the site, with frequent sampling
 (1x every 2weeks) in deciduous forest sites during senescence, and infrequent 
 year‐round sampling (1x every 1‐2 months) at evergreen sites.  Ground traps and
 elevated traps are paired
 2.A subset of collection bouts are chosen for chemistry and stable isotope 
 measurements, once every five years.
 3. Traps are placed within 4 40m x 40m tower plots and 26 20m x 20m plots.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r, dimensions of Neonics dataset}
summary_neonics <- summary(Neonics)
summary_neonics
dim_neonics <- dim(Neonics)
dim_neonics
ncol(Neonics)
nrow(Neonics)
#These various commands return the basic dimensions of the Neonics dataset, how 
 #many columns/rows are in the dataframe.
```

6.  Using the `summary` function on the "Effect" column, determine the most 
common effects that are studied. Why might these effects specifically be of 
interest?

```{r, common effects of Neonics}
Effect_Neo <- summary(Neonics$Effect)

Effect_Neo

#This commands allows us to see how many studies return results on the each 
#effect of Neonics.
```

> Answer:Population and Mortality. These 2 effects are of the greatest interest 
b/c they say the most about the lethality of neonicotinoids and their impact on 
insect populations. Population is a measure of abundance so if the abundance 
falls drastically the insect may be functionally extinct in the region. 


7.  Using the `summary` function, determine the six most commonly studied 
species in the dataset (common name). What do these species have in common, and 
why might they be of interest over other insects? Feel free to do a brief 
internet search for more information if needed.[TIP: The `sort()` command 
can sort the output of the summary command...]

```{r, Important insects impacted by Neonics}
six_insects <- summary(Neonics$Species.Common.Name)
six_insects
sort(six_insects,decreasing=FALSE)
#As with the summary of effects, these commands let us see which insects are of
#most interest by returning how many studies include each insect.
```

> Answer:Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, 
Bumble Bee, Italian Honeybee. Other than the parasitic wasp, all these insects 
have in common is that they're all bees and they're all important pollinators. 
They are important pollinators and the honey bees are very economically 
important for their pollinating and their honey. 

8.  Concentrations are always a numeric value. What is the class of 
`Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r, what class of data is Conc.1..Author}
lapply("Conc.1..Author",class)

#It's useful to know which class of data you're dealing with because this 
#informs the commands that can be used and what type of information it is
```

> Answer: It's a character class. There are non-numeric characters in the 
column, like / and NR. Concentrations were not reported for all rows.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r, frequency plot of Publication Years}
ggplot(Neonics,aes(Publication.Year))+
  geom_freqpoly(bins=500)
  
#Looking at frequency of each publication year in a frequency plot. 



```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r, Frequency Plot of Publication Year and test site}
ggplot(Neonics,aes(Publication.Year)) +
  geom_freqpoly(aes(color=Test.Location), bins=500)

#This allows us to look at multiple questions, in which years were most studies
#published and where were those studies conducted. 
```

Interpret this graph. What are the most common test locations, and do they 
differ over time?

> Answer:Field natural and Lab are the most common test locations and after
2003 Test location is overwhelmingly in the lab whereas before 2000 it's
mostly in field natural.

11. Create a bar graph of Endpoint counts. What are the two most common end 
points, and how are they defined? Consult the ECOTOX_CodeAppendix for more 
information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = , 0.5
hjust=1))` to the end of your plot command to rotate and align the
X-axis labels...]

```{r, Adjusting graph labels}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#This command simply makes the graph labels easier to read
```

> Answer:LOEL:Lowest-observable-effect-level: lowest dose (concentration) 
producing effects that were significantly different (as reported by authors) 
from responses of controls and NOEL:No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses 
of controls. So the most common endpoints were low doses producing effects and
high doses producing no effects.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date
and confirm the new class of the variable. Using the `unique` function, 
determine which dates litter was sampled in August 2018.

```{r,fixing date column}
lapply("collectDate",class)

Litter$collectDate <- "2018-08-02"
litter_dates <- ymd(Litter$collectDate)
litter_dates


lapply("collectDate",class)

collect_dates<- unique(Litter$collectDate,incomparables = FALSE)

#August 2, 2018 (and August 30, 2018 but I accidentally got rid of that)
#fixing the date columns b/c R is weird about dates
```

13. Using the `unique` function, determine how many plots were sampled at Niwot 
Ridge. How is the information obtained from `unique` different from that
obtained from `summary`?

```{r,unique vs. summary}
Niwot_Ridge_plots<- unique(Litter$plotID, incomparables=FALSE)
Niwot_Ridge_plots
Niwot_test <- summary(Litter$plotID)
Niwot_test

#Comparing unique and summary. As well as seeing how many plots were sampled.
```

> Answer: 12 plots were sampled at Niwot Ridge. Unique will return the same 
vector, dataframe, or array with duplicates removed so you can see how many 
exist in each row. Summary summarizes each unique plot, so returns how many 
subplots are in each plot, without returning the number of distinct plots. 

14. Create a bar graph of functionalGroup counts. This shows you what type 
of litter is collected at the Niwot Ridge sites. Notice that litter types 
are fairly equally distributed across the Niwot Ridge sites.

```{r, bar chart of litter types}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar()

#Looking at what types of litter were collected at each site. 

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot
of dryMass by functionalGroup.

```{r, boxplot vs. violin plot}
ggplot(Litter) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))


ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup), 
              draw_quantiles = c(0.25, 0.5, 0.75))

#Comparing how a boxplot and a violin plot displayed the same data.

```

Why is the boxplot a more effective visualization option than the violin plot 
in this case?

> Answer:Boxplot shows more of the summary statistics with mean and 1st and
2nd quartile and outliers distinct from the rest. The violin plot does not 
isolate the outliers so in this case it appears the Twigs/branches is much 
more dryMass than is the case.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles made up the largest amount of biomass, followed by Mixed
litter, with Twigs and branches in third.

