---
title: 'Assignment 9: Spatial Analysis in R'
author: "Amanda Sajewski"
output:
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single HTML file.


## DATA WRANGLING

### Set up your session

1. Import libraries: tidyverse, sf, leaflet, and mapview
2. Check your working directory

```{r Prepare workspace}
#1. 
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)

#2. 
getwd()

```

### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the
US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` 
(as North Carolina's was `37`).

3. Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, 
filtering records for Nebraska counties (State FIPS = 31)
4. Reveal the dataset's coordinate reference system
5. Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties

Nebraska_stream <- st_read("Data/Spatial/cb_2018_us_county_20m.shp")%>%
  filter(STATEFP==31)


#4. Reveal the CRS of the counties features

st_crs(Nebraska_stream)

#The coordinate reference system is NAD83 (epsg 4269)

#5. Plot the data

mapview(Nebraska_stream)

```

6. What is the EPSG code of the Counties dataset? Is this a geographic or a 
projected coordinate reference system? (Or, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: look the EPSG 
code on <https://spatialreference.org>) 

> ANSWER: 
The EPSG code of this dataset is 4269. It's a geographic coordinate reference 
system, using angular coordinate units. It's associated with the "North American 
Datum 1983"
> 


### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` 
folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7. Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe.

8. Display the column names of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe

USGS_gage_location <- read.csv("Data/Raw/NWIS_SiteInfo_NE_RAW.csv")


#8. Reveal the names of the columns

colnames(USGS_gage_location)

```

9. What columns in the dataset contain the x and y coordinate values, 
respectively?  
> ANSWER:
x coordinate values = "dec_long_va"
y coordinate values = "dec_lat_va"
> 


### Convert the dataframe to a spatial features ("sf") dataframe

10. Convert the dataframe to an sf dataframe.
 * Note: These data use the same coordinate reference system as the counties
 dataset

11. Display the column names of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object
USGS_gage_sf <- USGS_gage_location %>%
  st_as_sf(coords=c("dec_long_va","dec_lat_va"),crs=4269)
USGS_gage_sf$site_no <- as.factor(USGS_gage_sf$site_no)


#11. Re-examine the column names

colnames(USGS_gage_sf)


```

12. What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: 
The new geometry field was created to attach to a map. This was created 
partially from the "dec_long_va" and "dec_lat_va" fields which disappeared.
>


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
 * Be sure the datasets are displayed in different colors
 * Title your plot "NWIS Gage Locations in Nebraska"
 * Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features
Gage_NE <- ggplot()+
  geom_sf(data=Nebraska_stream)+
  geom_sf(data=USGS_gage_sf, aes("station_nm"='blue'))+
  labs(title="NWIS Gage Locations in Nebraska", subtitle = "Amanda Sajewski")

print(Gage_NE)



```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name 
and site number along with stream gage heights (in meters) recorded during the 
recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is 
found in the Data/Raw folder.

14. Read in and wrangle the `NWIS_SiteFlowData_NE_RAW.csv` dataset: 
  * Be sure the `site_no` column is read in as a factor
  * Group by `site_no` and `station_nm`, computing the mean `gage_ht` across 
  groups values
15. Show the column names .
16. Join our site information (already imported above) to these gage height data.
 * The `site_no` and `station_nm` can both/either serve as joining attributes.
 * Construct this join so that the result only includes spatial features where 
 both tables have data.
17. Show the column names in this resulting spatial features object
18. Show the dimensions of the resulting joined dataframe

```{r Combining gage height data with site location}
#14. Read the site flow data into a data frame

NWIS_site <- read.csv("Data/Raw/NWIS_SiteFlowData_NE_RAW.csv")
NWIS_site$site_no <- as.factor(NWIS_site$site_no)
class(NWIS_site$site_no)
#coverting it to a factor was necessary so the dataframes could be merged b/c
#the other site_no was a factor

NWIS_site <- NWIS_site %>%
  group_by(site_no,station_nm)%>%
  summarize(
    mean_gage_ht = mean(gage_ht, na.rm=T))

#15. Show the column names

colnames(NWIS_site)

#16. Join the flow data to our NWIS gage location spatial dataframe


NE_gage_flow <- merge(x =USGS_gage_sf, 
                      y=NWIS_site,
                      by.x="site_no",
                      by.y="site_no")
#merge was necessary b/c using left_join somehow made it so geometry was no 
#longer a location/spatial object

#17. Show the column names of the joined dataset

colnames(NE_gage_flow)

#18. Show the dimensions of this joined dataset

dim(NE_gage_flow)



```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage 
heights spatially.
19. Plot the gage sites on top of counties (using `mapview`, `ggplot`, or 
`leaflet`) 
 * Show the magnitude of gage height by color, shape, other visualization 
 technique.

```{r Mapping gage height}
#Map the points, sized by gage height


Gage_ht_NE <- ggplot()+
  geom_sf(data=Nebraska_stream)+
  geom_sf(data=NE_gage_flow, aes(size = mean_gage_ht,fill=mean_gage_ht),pch=21,alpha=.6)+
  guides(size=FALSE)+
  labs(title="NWIS Gage Heights in Nebraska", subtitle = "Amanda Sajewski",fill="Average gage height(m)")+
  xlab("Longitude") +
  ylab ("Latitude")
  
  
print(Gage_ht_NE)

```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we 
should transform our data into a projected coordinate system. We'll choose UTM 
Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as 
different colors

```{r Transforming datasets to UTM Zone 14N, Nebraska counties}
#20 Transform the counties and gage location datasets to UTM Zone 14

counties_utm <- st_transform(Nebraska_stream,crs=32614)
gage_ht_utm <- st_transform(NE_gage_flow,crs=32614)

#21 Plot the data

Gage_ht_NE_utm <- ggplot()+
  geom_sf(data=counties_utm)+
  geom_sf(data=gage_ht_utm, aes(size = mean_gage_ht,fill=mean_gage_ht),pch=22,col="red",alpha=.6)+
  guides(size=FALSE)+
  labs(title="NWIS Gage Heights in Nebraska", subtitle = "UTM Projection",fill="Average gage height(m)")+
  xlab("Longitude") +
  ylab ("Latitude")

print(Gage_ht_NE_utm)

#fill is used for the legend title, pch is the size/shape of the points


```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Douglas county from your projected county sf dataframe 
23. Select the gage sites falling `within` that county
 * Use either matrix subsetting or tidy filtering 
24. Create a plot showing:
 * all Nebraska counties,
 * the selected county, 
 * and the gage sites in that county

```{r Selecting Douglas County and the gage sites there}
#22 Select the county
Douglas <- counties_utm %>%
  filter(NAME=="Douglas")


#23 Select gages within the selected county
Douglas_gages <- gage_ht_utm[Douglas,]

#simple intersect

#24 Plot

Douglas_gage_site <- 
  mapview(counties_utm,
        col.regions = 'yellow', 
        alpha.regions = 0.4,
        map.types = "CartoDB.Positron",
        legend = FALSE)+
  mapview(Douglas,
          col.regions = "red",
          alpha.regions = .8)+
  mapview(Douglas_gages,
          col.regions = "blue",
          alpha.regions = .9,
          cex=2)

Douglas_gage_site

#alpha.regions determines the opacity of layers, map.types is the background map,
#cex is the size of the points


```
