---
title: "Assignment 2: Coding Basics"
author: "Amanda Sajewski"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r} 
#1.Created a sequence increasing by 4's from 1-100
seqby4 <-seq(1,100,4)

#2. Determined the mean of the sequence
mean(seqby4)

median(seqby4)

#3. Is mean or median greater
49>49


```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
names <- c("Sam", "Maria", "John", "Mark") 
#character vector
scores <-c(45,90,78,88)
#numeric vector
Passed <- c("False", "True","True", "True")
#character vector
df_test_scores<- as.data.frame(names)

df_testscores<-cbind(df_test_scores,scores,Passed)

#9.  QUESTION: How is this data frame different from a matrix?

 #Answer:This dataframe contains both numeric and character classes of vectors whereas matrices can only contain one.

#10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 


#11. Apply your function to the vector with test scores that you created in number 5.


grade<-function(x){ifelse(x>50,"Pass","Fail")} #requirement,if true, if false
passing_grade<-grade(scores) #apply ifelse statement to scores vector
passing_grade #printed results
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' and 'if' and 'else' are the same. Just ifelse is a neater line of code, so either would've worked.
